|

### 2024-05-09 Drafts/Paper_1/03-methods.Rmd
Methods results done

## Data

Data from Julie's repo, cite Julie. This data repository had sufficient data to run the analysis for 77 stocks from NE Atlantic, NW Atlantic, and the North Pacific, this includes stocks managed by Fisheries and Oceans Canada (DFO), NOAA, and ICES. See Appendix X for details on the stocks included. These stocks all have age-specific estimates of abundance, biomass, removals, natural mortality, and maturity.

## Incidental Catch Model (ICM)
## Reconstrutive Life-Table Analysis (REC)
## Life Table Reconstruction (LITAR)

@@

The discrete time Euler-Lotka equation can be shown to be the characteristic polynomial of the Leslie Matrix, where $f_0$ is the fecundity of the first age class with mature individuals [@leslieUseMatricesCertain1945,@caswellMatrixPopulationModels2001]:

$$

  \begin{bmatrix}
    f_0  & f_1 & f_2  & f_3    & ...          & f_{\omega-1} \\
    s_0  & 0   & 0    & 0      & ...          & 0 \\
    0    & s_1 & 0    & 0      & ...          & 0 \\
    0    & 0   & s_2  & 0      & ...          & 0 \\
    0    & 0   & 0    & \ddots & ...          & 0 \\
    0    & 0   & 0    & ...    & s_{\omega-2} & 0 \\
  \end{bmatrix}
$$

@@

The Euler-lotka equation requires the fecundity given as the number of offspring produced by each female in each age class. This can be estimated from the available recruitment, age at maturity, and biomass information. **Note there is no need to divide by 2 for females because we'd divide the recruits by 2 to get the number of females, and the spawning numbers by 2, which would just cancel out**.

@@

$$
RPA_{y,a} =  SSB_{(y-ra),a} RPS_{y,a}
$$
This method assumes that there is no effect of age on reproductive success (i.e., each kilogram of spawning biomass contributes the same to the total total reproductive output, irrespective of their age). Finally, the fecundity ($Fec_{y-1,a}$) is calculated as number of offspring produced by each individual in the population, this is offset by one year, as the recruits observed in year $y+1$ are produced in the Euler-Lotka formulation in year $y$.

$$
Fec_{y,a} =  RPA_{y+1,a} / SSN_{(y-ra+1),a}
$$
Due to challenges with separability in typical stock assessment models, natural mortality is often assumed for each age class, in some assessments it varies by age and over time, in other assessments it does not.  Irrespective of this, the stock assessment used in this analysis all provide an estimate of natural mortality by age and year, this value is typically an instantaneous rate and is convert to a proportional natural mortality for the Euler-Lotka model.

@@

$$
FM_{y,a} =  Catch_{y,a} / N_{y,a}
$$
This is combined with the natural mortality to get a total mortality estimate ($Z_{y,a}$) for each age in each year

@@

$$
N_{y+1} = \lambda{y} N_{y}
$$
The result of this was compared to the abundance estimate from the stock assessment for the upcoming year ($N_{y+1}$). Where necessary, the fecundity and natural mortality estimates were adjusted until the $\lambda$ estimate from the Euler-Lotka equation resulted in an abundance time-series estimate ($N_{y+1}$) that was less than 5% different from the abundance estimates from the stock assessment (AVERAGE DIFFERNCE WAS 0.XX%). In this way the $\lambda$, fecundity, and natural mortality estimates are obtained for each year in which data are available.

WONDER IF I HAVE THE "AVERAGE" DIFFERENCE of NM and FECUNDITY FROM THE ORIGINAL VALUES FROM THE ABOVE or is that opening up pandora's box.

@@

Doubling time was calculated using a simulation approach in which the estimated fecundity and natural mortality for a year were sampled from the historically observed fecundity and natural natural mortality. The Euler-Lotka equation was used to estimate the population growth rate and this rate was used to adjust the population biomass accordingly.  The simulation was run for XX years with YY simulations, the percentage of the simulations in which the population size had doubled was recorded each year. If a population had doubled in size, it was removed from the simulation and recorded as having doubled for all subsequent years.

@@

The lifetime reproductive success ($R_0$) in the absence of fishing was calculated for each cohort in which fecundity and natural mortality were available for all age classes. This was calculated for each cohort as

@@

The generation time here is defined as the mean age of the spawners of the observed recruits in a cohort over its lifetime ($T_G$) in the absence of fishing was calculated for each cohort in which fecundity and natural mortality were available for all age classes. This was calculated for each cohort as.

@@

>>>>

**NOTE THAT Walleye Pollock was T. chalcogramma and is now G. chalcogrammus**
## Data

Data from Julie's repo, cite Julie. This data repository had sufficient data to run the analysis for 77 stocks from NE Atlantic, NW Atlantic, and the North Pacific, this includes stocks managed by Fisheries and Oceans Canada (DFO), NOAA, and ICES. See Table \@ref(tab:tab-meta) for details on the stocks included. The stocks included in this analysis all have age-specific estimates of abundance, biomass, removals, natural mortality, and maturity.

## Incidental Catch Model (ICM), Reconstrutive Life-Table Analysis (REC), Life Table Reconstruction (LITAR)

@@

The discrete time Euler-Lotka equation can be shown to be the characteristic polynomial of the Leslie Matrix, where $f_0$ is the fecundity of the first age class with mature individuals [@leslieUseMatricesCertain1945, @caswellMatrixPopulationModels2001]:

\begin{align}
  \begin{bmatrix}
    f_0  & f_1 & f_2  & f_3    & ...          & f_{\omega-1} \\
    s_0  & 0   & 0    & 0      & ...          & 0 \\
    0    & s_1 & 0    & 0      & ...          & 0 \\
    0    & 0   & s_2  & 0      & ...          & 0 \\
    0    & 0   & 0    & \ddots & ...          & 0 \\
    0    & 0   & 0    & ...    & s_{\omega-2} & 0 \\
  \end{bmatrix}
\end{align}

@@

The Euler-lotka equation requires the fecundity given as the number of offspring produced by each female in each age class. This can be estimated from the available recruitment, age at maturity, and biomass information. **Delete this, but Note there is no need to divide by 2 for females because we'd divide the recruits by 2 to get the number of females, and the spawning numbers by 2, which would just cancel out**.

@@

$$
RPA_{y,a} =  SSB_{(y-ra),a} RPS_{y,a}
$$
This method assumes that there is no effect of age on reproductive success (i.e., each kilogram of spawning biomass contributes the same to the total total reproductive output, irrespective of their age). Finally, the fecundity ($Fec_{y,a}$) is calculated as number of offspring produced by each individual in the population, this is offset by one year, as the recruits observed in year $y+1$ enter into the Euler-Lotka formulation in year $y$.

$$
Fec_{y,a} =  \frac{RPA_{y+1,a}}{SSN_{(y-ra+1),a}}
$$
Due to challenges with separability in typical stock assessment models, natural mortality is often assumed for each age class, in some assessments it varies by age and over time, in other assessments it does not.  Irrespective of this, the stock assessment used in this analysis all provide an estimate of natural mortality by age and year, this value is typically an instantaneous rate and is convert to a proportional natural mortality for the Euler-Lotka model.

@@

$$
FM_{y,a} =  \frac{Catch_{y,a}}{N_{y,a}}
$$
This is combined with the natural mortality to get a total mortality estimate ($Z_{y,a}$) for each age in each year

@@

$$
N_{y+1} = \lambda{y} N_{y}
$$
The result of this was compared to the abundance estimate from the stock assessment for the upcoming year ($N_{y+1}$). Where necessary, the fecundity and natural mortality estimates were adjusted until the $\lambda$ estimate from the Euler-Lotka equation resulted in an abundance time-series estimate ($N_{y+1}$) that was within 5% of the abundance estimates from the stock assessment (median difference was `r med.per.diff`). In this way the $\lambda$, fecundity, and natural mortality estimates are obtained for each year in which data are available. The realized $\lambda_{F>0}$ included fishing mortality, in addition $\lambda_{F=0}$ could also be calculated by using annual the natural mortality and fecundity estimates. $\lambda_{F=0}$ provides an estimate of the potential population growth in the absence of fishing in a give year.

@@

Doubling time in the absence of fishing was calculated using a simulation approach in which the estimated fecundity and natural mortality for a projection year were sampled from the fecundity and natural natural mortality estimated in each year for each stock. The Euler-Lotka equation was used to estimate the population growth rate and this rate was used to adjust the population biomass accordingly.  The simulation was run for `n.sim.years` years with `r n.dt.sims` simulations, the percentage of the simulations in which the population size had doubled was recorded each year. If a population had doubled in size, it was removed from the simulation and recorded as having doubled for all subsequent years. This approach implicitly assumes no density dependence effects.

@@

The lifetime reproductive success ($R_0$)  was calculated for each cohort in which fecundity and natural mortality were available for all age classes. It was calculated using only natural mortality (F=0) and with the combined effect of natural mortality and fishing mortality (F>0).  This was calculated for each cohort as

@@

The generation time was defined as the mean age of the spawners of the observed recruits in a cohort over its lifetime ($T_G$), here we calculated this for each cohort in which fecundity and natural mortality were available for all age classes. It was calculated twice, using only natural mortality (F=0) and with the combined effect of natural mortality and fishing mortality (F>0).

@@

## Density Dependence

Density dependence was explored by modelling the estimated $\lambda$ against the proportion of the maximum abundance observed for each stock.  This was done twice, once using ($\lambda_{F=0}$) and once using ($\lambda_{F>0}$).  Generalized Additive Models were used to allow for some non-linearity in the relationships (CITE MCGV and WOOD), the overall effect and the effect for each stock was modelled as

\begin{align}
\lambda_{i,j} \sim N(0,\sigma^2_{i,j}) \\
E(\lambda_{i,j}) = \mu_{i,j} \\
log(\mu_{i,j}) =  Prop_{i} + Stock_{j}  \times Prop_{i} \\
Stocks_{i,j} \sim N(0,\sigma^2_{Stocks})
\end{align}

The interaction was modeled using the *fs* basis using the mgcv package in R, this treats the smooth curves as random effect and is used when a factor has a large number of levels (CITE!).  Various alternative modelling methods were explored (various model formulations, alternative error distributions, etc.) but all models had similar challenges overcoming either the underlying non-normality or the heteroskedasticity of the data. As such, the uncertainty surrounding the model output should be treated with caution, while we discuss the general trends this model provides we attempted to avoid discussion of the statistical significance of the trends.

### 2024-05-09 Drafts/Paper_1/05-discussion.Rmd
Methods results done

Looks like density dependence is real. Dave's pet theroy is there is solid evidence of density dependence for many stocks above around 0.4, below 0.4 the evidence hints more at density independence for many stocks, BUT, some stocks have an ability to very occasionally very spectacularly take advantage of the environment.  But not all stocks and species have this ability and the strength and frequency seems to be influenced by where you life and what species you are. Little evidence for spectacular reproduction events in NW Atlantic, while the majority of species assessed in the north sea have this ability. Cod show this dicomaty perfectly....

@@

>>>>

Looks like density dependence is real. Dave's pet theory is there is solid evidence of density dependence for many stocks above around 0.4, below 0.4 the evidence suggest that density dependence weakens for some stocks, BUT, some stocks have an ability to very occasionally very spectacularly take advantage of the environment.  But not all stocks and species have this ability and the strength and frequency seems to be influenced by where you live and what species you are. Little evidence for high lambda for most stocks in the NW Atlantic, while many stocks assessed in the North Sea had anomalously high lambdas in 1 or more years. Cod show this  perfectly....

@@

By ignoring the density dependence patterns observed in this analysis, the doubling time analysis would overestimate the doubling time for stocks that were at low abundance with classic negative density dependence dynamics, while it would underestimate it for stocks with positive density dependence. Here perhaps talk about infrequent big events and how using a theoretical rmax within a logistic model isn't a good idea because this will way over-estimate recovery potential at low abundance.

The optimization procedure assumes that both the estimates of abundance and the fishery removals from the stock assessment are correct. *A couple sentences about why fisheries removals may be underestimated*. This would result in the estimates of the $\lambda$ excluding fishing mortality in this analysis being lower than reality, as this 'missing' mortality was accounted for by reductions in either fecundity or an increase in natural mortality. The optimization routine resulted in a median decrease in the no-fishing $\lambda$ estimates of approximately `r med.lam.per.diff`%, the overall directionality of this change is consistent with fishing mortality being generally underestimated.

Fisheries appear to slightly weaken the effects of density dependence overall. Maybe due to fisheries having positive density dependence? Seemingly below around 0.5 fishing leads to a positive lambda, below 0.5 it is a negative lambda... that's kinda kooky cool!

The stocks which reached very low abundance (< 10% of maximum) tended to have high values of lambda, suggesting strong compensation in stocks when they are at extremely low abundance. Oddly, the path to this isn't linear, and positive density dependence (or at least density independence was evident between 10 and 40% of maximum abundance).

Probably something worth citing in here @thorsonPredictingRecruitmentDensity2020

### 2024-05-09 Drafts/Paper_1/08-figures.Rmd
Methods results done

>>>>

# Figures {-}

```{r lambda-bx-6-fig, echo=F, fig.cap="$\\lambda$ estimates for each stock of G. morhua, M. aeglefinus, P. virens, C. harengus, P. platessa, and S. solea."}
knitr::include_graphics(paste0(loc,"/Figures/tuned/No_age_correction/Lambda_boxplot_6_species_",what.2.tune,".png"))

\newpage
```{r lambda-bx-other-fig, echo=F, fig.cap="$\\lambda$ estimates for the remainded of the stocks in this analysis, plotted by the location of the stock. The abreviated species name is provided on the x axis."}
knitr::include_graphics(paste0(loc,"/Figures/tuned/No_age_correction/Lambda_boxplot_other_species_",what.2.tune,".png"))

\newpage
```{r lambda-sd-fig, echo=F, fig.cap="Standard deviation of the $\\lambda$ estimates for each stock, ordered by species, in this analysis. "}
knitr::include_graphics(paste0(loc,"/Figures/tuned/No_age_correction/Lambda_sd_",what.2.tune,".png"))
```

\newpage
```{r dd-overall-plt, echo=F, fig.cap="Density dependence of $\\lambda$ The top panel shows $\\lambda_{F=0}$ against the proportion of maximum abundance, with a GAM smooth and the GAM estimated 95\\% confidence interval. The middle panel shows $\\lambda_{F>0}$ against the proportion of maximum abundance, with a GAM smooth and the GAM estimated 95% confidence interval. The bottom panel compares the GAM smooths for $\\lambda_{F=0}$ and $\\lambda_{F>0}$."}
knitr::include_graphics(paste0(loc,"/Figures/tuned/No_age_correction/density_dependence_overall_",what.2.tune,".png"))
```

\newpage
```{r dd-nf-stock-plt, echo=F, fig.cap="Density dependence of $\\lambda_{F=0}$ for each stock in the analysis plotted regionally."}
knitr::include_graphics(paste0(loc,"/Figures/tuned/No_age_correction/density_dependence_nf_stocks_",what.2.tune,".png"))
```

\newpage
```{r dd-fish-stock-plt, echo=F, fig.cap="Density dependence of $\\lambda_{F>0}$ for each stock in the analysis plotted regionally."}
knitr::include_graphics(paste0(loc,"/Figures/tuned/No_age_correction/density_dependence_fish_stocks_",what.2.tune,".png"))
```

\newpage
```{r r0-6-plt, echo=F, fig.cap="$R_0$ estimates for each stock of G. morhua, M. aeglefinus, P. virens, C. harengus, P. platessa, and S. solea."}
knitr::include_graphics(paste0(loc,"/Figures/tuned/No_age_correction/R0_boxplot_6_species_",what.2.tune,".png"))
```

\newpage
```{r r0-other-plt, echo=F, fig.cap="$R_0$ estimates for each stock of G. morhua, M. aeglefinus, P. virens, C. harengus, P. platessa, and S. solea."}
knitr::include_graphics(paste0(loc,"/Figures/tuned/No_age_correction/R0_boxplot_other_species_",what.2.tune,".png"))
```

\newpage
```{r r0-nf-ts-plt, echo=F, fig.cap="$R_0$ estimates for every cohort for each species excluding the effects of fishing."}
knitr::include_graphics(paste0(loc,"/Figures/tuned/No_age_correction/r0_heatmap_nf_",what.2.tune,".png"))
```

\newpage
```{r r0-fish-ts-plt, echo=F, fig.cap="$R_0$ estimates for every cohort for each species including the effects of fishing."}
knitr::include_graphics(paste0(loc,"/Figures/tuned/No_age_correction/r0_heatmap_fish_",what.2.tune,".png"))
```

\newpage
```{r gl-6-plt, echo=F, fig.cap="Generation length estimates for each stock of G. morhua, M. aeglefinus, P. virens, C. harengus, P. platessa, and S. solea."}
knitr::include_graphics(paste0(loc,"/Figures/tuned/No_age_correction/GL_boxplot_6_species_",what.2.tune,".png"))
```

\newpage
```{r gl-other-plt, echo=F, fig.cap="Generation length estimates for each stock of G. morhua, M. aeglefinus, P. virens, C. harengus, P. platessa, and S. solea."}
knitr::include_graphics(paste0(loc,"/Figures/tuned/No_age_correction/GL_boxplot_other_species_",what.2.tune,".png"))
```

\newpage
```{r gl-nf-ts-plt, echo=F, fig.cap="Generation length estimates for every cohort for each species excluding the effects of fishing."}
knitr::include_graphics(paste0(loc,"/Figures/tuned/No_age_correction/GL_heatmap_nf_",what.2.tune,".png"))
```

\newpage
```{r gl-fish-ts-plt, echo=F, fig.cap="Generation length estimates for every cohort for each species including the effects of fishing."}
knitr::include_graphics(paste0(loc,"/Figures/tuned/No_age_correction/GL_heatmap_fish_",what.2.tune,".png"))
```

<!-- Now for the Doubling time -->

\newpage
```{r dub-time-map, echo=F, fig.cap="Doubling time estimate for each stock with no fishing mortality."}
knitr::include_graphics(paste0(loc,"/Figures/tuned/No_age_correction/Doubling_time_heatmap_by_order_",what.2.tune,".png"))
```

\newpage
```{r dub-ts, echo=F, fig.cap="The number of stocks which doubled in size by a given year with no fishing mortality  The dashed line represents the 77 stocks in the analysis."}
knitr::include_graphics(paste0(loc,"/Figures/tuned/No_age_correction/Doubling_time_time_series_",what.2.tune,".png"))
```

\newpage
```{r dub-vs-sd, echo=F, fig.cap="The number of stocks which doubled in size by a given year with no fishing mortality  The dashed line represents the 77 stocks in the analysis."}
knitr::include_graphics(paste0(loc,"/Figures/tuned/No_age_correction/Doubling_time_sd_effects_",what.2.tune,".png"))
```

# Supplemental Figures {-}

### 2024-05-09 Drafts/Paper_1/index.Rmd
Methods results done

```{r load-libraries-n-data, echo = FALSE, cache = FALSE, message = FALSE, results = 'hide', warning = FALSE}
# add other packages here:
library(dplyr)
library(readr)
library(tibble)
library(csasdown)
# OK, so using the ICES assessments here's what we get for North Sea cod.
library(readxl)
library(tidyverse)
library(rio)
library(ggthemes)
library(cowplot)
library(gtable)
library(viridis)
theme_set(theme_few(base_size = 14))

# The location of your data
loc <- 'D:/GitHub/ICM'

# If we already have the results don't re-run the models or load data we don't need.
load.results <- T
# What tunning model are you running or did you run!
what.2.tune <- 'fec_nm'

>>>>

```{r load-libraries-n-data, echo = FALSE, cache = FALSE, message = FALSE, results = 'hide', warning = FALSE}
# add other packages here:
library(dplyr)
library(readr)
library(tibble)
library(csasdown)
# OK, so using the ICES assessments here's what we get for North Sea cod.
library(readxl)
library(tidyverse)
library(rio)
library(ggthemes)
library(cowplot)
library(gtable)
library(viridis)
library(arm)
library(mgcv)
library(lme4)
library(scales)
library(ggh4x)
library(gratia)
theme_set(theme_few(base_size = 14))

# The location of your data
loc <- 'D:/GitHub/ICM'
# The parameters for the doubling time simulation
n.dt.sims <- 1000 # Increase to 1000 for final run, which takes like a full day or something 
n.sim.years <- 100
# If we already have the results don't re-run the models or load data we don't need.
load.results <- T
# What tunning model are you running or did you run!
what.2.tune <- 'fec_nm'
